---
title: "Tandem repeat analysis in Darevskia nairensis and D. valentini"
author: "Sviatoslav Sidorov, Pavel Nikitin"
date: "11/11/2020"
output: html_document
---

## Already found probes
V1 - 5′ GCCCCACAAATAACCCAGAGATGCATTTTA
V2 - 5′ TGTCACGCGTCCATGGAGATCCACGTGATG

N1 - 5' AGAAAATCAGGCCTCGGCAACTACACCTGC
N2 - 5' CAGGTCGAAATGCATAGAAAACACACCAGC
N3 - 5' GACGTACTTGGCTTTTGTAACGTTCAATTT

```{r setup, include=TRUE}
require("knitr")
opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(echo = TRUE)
stringsAsFactors = F

library(tidyverse)
library(kableExtra)
library(utils)
```

## Filter repeat units

Read in the tables with the tandem repeat units:

```{r, include=T}
tr.units.n = read.delim("../data/input/contigs/N_top10pc_tab_bycol.tsv", header = F, sep = "\t", stringsAsFactors = F) 
names(tr.units.n) = c("contig", "length", "mean.coverage", "repr", "unit.seq")

tr.units.v = read.delim("../data/input/contigs/V_top10pc_tab_bycol.tsv", header = F, sep = "\t", stringsAsFactors = F)
names(tr.units.v) = c("contig", "length", "mean.coverage", "repr", "unit.seq")
```

Filter the repeat units:

```{r, include=T}
# Use the following commented code when removing probes that have polyA/C/G/T runs of length >=6
  # dplyr::filter(!stringr::str_detect(unit.seq, "AAAAAA")) %>%
  # dplyr::filter(!stringr::str_detect(unit.seq, "TTTTTT")) %>%
  # dplyr::filter(!stringr::str_detect(unit.seq, "GGGGGG")) %>%
  # dplyr::filter(!stringr::str_detect(unit.seq, "CCCCCC")) %>%
  
tr.units.n = tr.units.n %>%
  dplyr::arrange(unit.seq) %>%
  dplyr::filter(nchar(unit.seq) > 6) %>% # Do this filtering further down instead (on the level of probes)
  group_by(unit.seq) %>%
  do(dplyr::arrange(., desc(.$mean.coverage))) %>%
  do(head(., 1)) %>%
  ungroup()

tr.units.v = tr.units.v %>%
  dplyr::arrange(unit.seq) %>%
  dplyr::filter(nchar(unit.seq) > 6) %>% # Do this filtering further down instead (on the level of probes)
  group_by(unit.seq) %>%
  do(dplyr::arrange(., desc(.$mean.coverage))) %>%
  do(head(., 1)) %>%
  ungroup()
```

## Find the most species-specific repeat units

Calculate the Levenshteine distance between all K-bp substrings from the repeat units of _D. nairensis_ and _D. valentini_:

```{r, include=T}
probe.length = 30

take_all_substrings = function(s, k) {
  if (nchar(s) < k) {
    s = paste0(rep(s, k %/% nchar(s) + 1), collapse = '')
  }
  return(unlist(purrr::map(1:(nchar(s) - k + 1), 
                             function(i) stringr::str_sub(s, i, i + k - 1))))
}

generate_substring_df = function(x, k) {
  s = x[1, "unit.seq"]
  df = data.frame(probe.seq = take_all_substrings(s, k),
                  stringsAsFactors = FALSE)
  df$contig = x$contig
  df$length = x$length
  df$mean.coverage = x$mean.coverage
  df$repr = x$repr
  return(df)
}

rm_microsatellites = function(probe.seq) {
  for(unit.length in 2:6) {
    unit.num = 6 - unit.length + 2
    coord.last = probe.length - unit.num * unit.length + 1
    delta.seq = seq(0, unit.length * (unit.num - 1), by = unit.length)
    for(coord in 1:coord.last) {
      unique.units = unique(
        unlist(
        lapply(delta.seq, function(delta) {
        stringr::str_sub(probe.seq, coord + delta, coord + delta + unit.length - 1)
      })))
      if (length(unique.units) == 1) {
        return(FALSE)
      }
    }
  }
  return(TRUE)
}

# Generating probes
tr.probes.n = tr.units.n %>%
  group_by(contig, unit.seq) %>%
  do(generate_substring_df(., probe.length)) %>%
  ungroup() %>%
  distinct() %>%
  group_by(probe.seq, contig) %>%
  dplyr::filter(!stringr::str_detect(probe.seq, "AAAAAA")) %>%
  dplyr::filter(!stringr::str_detect(probe.seq, "TTTTTT")) %>%
  dplyr::filter(!stringr::str_detect(probe.seq, "GGGGGG")) %>%
  dplyr::filter(!stringr::str_detect(probe.seq, "CCCCCC")) %>%
  dplyr::filter(rm_microsatellites(probe.seq)) %>%
  ungroup()

tr.probes.v = tr.units.v %>%
  group_by(contig, unit.seq) %>%
  do(generate_substring_df(., probe.length)) %>%
  ungroup() %>%
  distinct() %>%
  group_by(probe.seq, contig) %>%
  dplyr::filter(!stringr::str_detect(probe.seq, "AAAAAA")) %>%
  dplyr::filter(!stringr::str_detect(probe.seq, "TTTTTT")) %>%
  dplyr::filter(!stringr::str_detect(probe.seq, "GGGGGG")) %>%
  dplyr::filter(!stringr::str_detect(probe.seq, "CCCCCC")) %>%
  dplyr::filter(rm_microsatellites(probe.seq)) %>%
  ungroup()

write.table(tr.probes.n, file="../data/results/tr_probes/tr_probes_n.tsv", quote = FALSE, sep = "\t", row.names = FALSE)
write.table(tr.probes.v, file="../data/results/tr_probes/tr_probes_v.tsv", quote = FALSE, sep = "\t", row.names = FALSE)

# Generate tables with all possible probes from N (done) and V
# From each repeat unit, take only distinct probes (done)
# Filter out probes that fully or partially consist of microsatellites or have polyA/C/G/T runs of length >=6 (done)
# Calculate Levenshtein distance for the remaining probes pairwise 

# Reshape the following commented code to calculate Levenshteine distance between probes instead of the whole repeats

# names(tr.probes.n) = c("unit.seq.n", "probe.seq.n", "contig.n", "length.n", "mean.coverage.n", "repr.n")
# names(tr.probes.v) = c("unit.seq.v", "probe.seq.v","contig.v", "length.v", "mean.coverage.v", "repr.v")
# tr.probes.n$dummy.col = 1
# tr.probes.v$dummy.col = 1
# tr.probes.all = tr.probes.n %>%
#   inner_join(tr.probes.v, by = ("dummy.col" = "dummy.col")) %>%
#   group_by(contig.n, probe.seq.n, contig.v, probe.seq.v) %>%
#   dplyr::mutate(edit.dist = utils::adist(probe.seq.n, probe.seq.v)) %>%
#   ungroup() %>%
#   dplyr::select(-dummy.col)
```

```{r, include=T}
tr.probes.n = read.delim("../data/results/tr_probes/tr_probes_n.tsv", header = T, sep = "\t", stringsAsFactors = F) 
tr.probes.v = read.delim("../data/results/tr_probes/tr_probes_v.tsv", header = T, sep = "\t", stringsAsFactors = F)

names(tr.probes.n) = c("unit.seq.n", "probe.seq.n", "contig.n", "length.n", "mean.coverage.n", "repr.n")
names(tr.probes.v) = c("unit.seq.v", "probe.seq.v","contig.v", "length.v", "mean.coverage.v", "repr.v")

tr.probes.v %>%
  dplyr::filter(contig.v == "CL4Contig833" & probe.seq.v == "AGGGAATATCAGCTCATTGATTCTGCTTCA")

tr.probes.n$dummy.col = 1
tr.probes.v$dummy.col = 1
tr.probes.all = tr.probes.n %>%
  inner_join(tr.probes.v, by = ("dummy.col" = "dummy.col")) %>%
  #dplyr::filter(contig.n == "CL1019Contig1" & probe.seq.n == "ACTAACAGAGTAGACTAACAGAGTAGACTA" & contig.v == "CL1026Contig1" & probe.seq.v == "ACTCGAGGGGCCAATGTGGCCGACTCGAGG")
  #dplyr::filter(contig.n == "CL1019Contig1" & probe.seq.n == "ACTAACAGAGTAGACTAACAGAGTAGACTA" & contig.v == "CL4Contig833" & probe.seq.v == "AGGGAATATCAGCTCATTGATTCTGCTTCA")
  
  # group_by(unit.seq.n, contig.n, probe.seq.n, unit.seq.v, contig.v, probe.seq.v) %>%%>%
  # dplyr::mutate(edit.dist = utils::adist(probe.seq.n, probe.seq.v)) %>%
  # ungroup() %>%
  # dplyr::select(-dummy.col)
```

Creating test data set while the script is running
```{r, include=T}
tr.probes.n = read.delim("../data/results/tr_probes/tr_probes_n.tsv", header = T, sep = "\t", stringsAsFactors = F)
tr.probes.v = read.delim("../data/results/tr_probes/tr_probes_v.tsv", header = T, sep = "\t", stringsAsFactors = F)

tr.probes.head.n = head(tr.probes.n, 50)
tr.probes.head.v = head(tr.probes.v, 50)

names(tr.probes.head.n) = c("unit.seq.n", "probe.seq.n", "contig.n", "length.n", "mean.coverage.n", "repr.n")
names(tr.probes.head.v) = c("unit.seq.v", "probe.seq.v","contig.v", "length.v", "mean.coverage.v", "repr.v")

tr.probes.head.n$dummy.col = 1
tr.probes.head.v$dummy.col = 1
tr.probes.head.all = tr.probes.head.n %>%
  inner_join(tr.probes.head.v, by = ("dummy.col" = "dummy.col")) %>%
  group_by(unit.seq.n, contig.n, probe.seq.n, unit.seq.v, contig.v, probe.seq.v) %>%
  dplyr::mutate(edit.dist = utils::adist(probe.seq.n, probe.seq.v)) %>%
  ungroup() %>%
  dplyr::select(-dummy.col)

#Analysing test data
tr.probes.head.min.all = tr.probes.head.all %>%
  group_by(probe.seq.n) %>%
  mutate(min.dist.n = min(edit.dist)) %>%
  ungroup() %>%
  group_by(probe.seq.v) %>%
  mutate(min.dist.v = min(edit.dist)) %>%
  ungroup() %>%
  group_by(probe.seq.n) %>%
  mutate(min.n = (edit.dist == min.dist.n)) %>%
  ungroup() %>%
  group_by(probe.seq.v) %>%
  mutate(min.v = (edit.dist == min.dist.v)) %>%
  ungroup()

#For D Nairensis
tr.probes.n.arranged = tr.probes.head.min.all %>%
  filter(min.n) %>%
  dplyr::arrange(desc(edit.dist))

write.table(tr.probes.n.arranged,
            "../data/results/N_tr_probes_by_distance_test.tsv",
            quote = F,
            sep = "\t",
            row.names = F)

kable(tr.probes.n.arranged %>%
  head(10)) %>%
kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(width = "100%")
```

```{r, include=T}
tr.probes.dist.geq10.n = read.delim("../data/results/tr_probes_all_min_dist_N.tsv", header = T, sep = "\t", stringsAsFactors = F) %>%
  dplyr::arrange(desc(mean.coverage.n))
tr.probes.dist.geq10.v = read.delim("../data/results/tr_probes_all_min_dist_V.tsv", header = T, sep = "\t", stringsAsFactors = F)%>%
  dplyr::arrange(desc(mean.coverage.v))

# tr.probes.dist.geq10.n %>%
#   dplyr::filter(probe.seq.n == 'GACGTACTTGGCTTTTGTAACGTTCAATTT')
```

¡Haven't yet used the code below:

Generate tables of repeat units, arranged by distance, for _D. nairensis_ and for _D. valentini_.

For _D. nairensis_:

```{r, include=T}
tr.units.n.arranged = tr.units.all %>%
  filter(min.n) %>%
  dplyr::select(contig.n, mean.coverage.n, unit.seq.n, edit.dist) %>%
  dplyr::arrange(desc(edit.dist)) %>%
  distinct()

write.table(tr.units.n.arranged,
            "data/results/N_tr_units_by_distance.tsv",
            quote = F,
            sep = "\t",
            row.names = F)

kable(tr.units.n.arranged %>%
  head(10)) %>%
kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(width = "100%")
```

For _D. valentini_:

```{r, include=T}
tr.units.v.arranged = tr.units.all %>%
  filter(min.v) %>%
  dplyr::select(contig.v, mean.coverage.v, unit.seq.v, edit.dist) %>%
  dplyr::arrange(desc(edit.dist)) %>%
  distinct()

write.table(tr.units.v.arranged,
            "data/results/V_tr_units_by_distance.tsv",
            quote = F,
            sep = "\t",
            row.names = F)

kable(tr.units.v.arranged %>%
  head(10)) %>%
kable_styling(bootstrap_options = c("striped", "hover")) %>%
  scroll_box(width = "100%")
```

```{r, include=T}
```